A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-20 paper.Rmd
done B-P modelling

@@

Is the sum of two components, *external* and *internal* causes. According to the *B-P*
theory, the external component is equal to the postneonatal mortality, which contains 
only an external component.

@@

and then fit a piecewise constant hazards model with cutpoints at `r gneocuts`, see 
Figure \@ref(fig:pchneo).

@@

### Putting all together

Looking at the full infant mortality interval on the $g$-scale, we have a piecewise 
constant hazards model with cutpoints at $g(0, 1, 2, 4, 7, 11, 16, 22, 28)$, see 
Figure \@ref(fig:together) (where the value close to zero is truncated).

@@

Now we would like to see the content of Figure \@ref(fig:together) on the natural time 
scale. If $Y = g(T)$, we have the relation between their respective hazard functions

@@

```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale."}
y <- hazards(fitall, cum = FALSE, n.points = 3000)$x
hy <- hazards(fitall, cum = FALSE, n.points = 3000)$y
te <- ginv(y)
hte <- dg(te) * hy
plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
     ylim = c(0, 0.002)) 
     ##xlim = c(0, 28))
axis(1, at = ginv(pcuts))
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

# Data

## Infant mortality, summary graphs

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950."}
infdat <- readRDS("data/infdat3_year3.rds")
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "IMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
abline(h = 0)
box()
```

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "NMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
abline(h = 0)
box()
```

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950."}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "PNMR")
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
y <- trend$event / trend$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "IMR")
abline(h = 0)
```

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "NMR")
abline(h = 0)
```

## Temperature

@@

```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950."}
oldpar <- par(mfrow= c(1, 2))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)")
text(17, 27, "Umeå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, uminavg, type = "b", col = "blue")
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)")
text(17, 27, "Skellefteå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, bminavg, type = "b", col = "blue")
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
##
par(oldpar)
```

>>>>

```{r excum, echo = FALSE}
fitex <- pchreg(Surv(enter, exit, event) ~ 1, cuts = g(c(28, 365)), data = ginfdat)
```

@@

Neonatal mortality is, according to Bourgeois-Pichat, the sum of two components, *external* and *internal* causes. According to the *B-P*
theory, the external component is equal to the postneonatal mortality, which contains 
only an external component.

@@

and then fit a piecewise constant hazards model with cutpoints at g(`r neocuts`), see 
Figure \@ref(fig:pchneo).

@@

### Putting it all together

Looking at the full infant mortality interval on the $g$-scale, we have a piecewise 
constant hazards model with cutpoints at $g(0, 2, 5, 8, 12, 17, 22, 25, 28, 365)$, see 
Figure \@ref(fig:together).

@@

<!--
Now we would like to see the content of Figure \@ref(fig:together) on the natural time 
scale. If $Y = g(T)$, we have the relation between their respective hazard functions

@@

```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale.", eval = FALSE}
y <- hazards(fitall, cum = FALSE, n.points = 3000)$x
hy <- hazards(fitall, cum = FALSE, n.points = 3000)$y
te <- ginv(y)
hte <- dg(te) * hy
plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
     ##ylim = c(0, 0.002)) 
     xlim = c(0, 28))
axis(1, at = pcuts)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

-->

The way to disentangle the endogenous and exogenous componets is to start with a plot 
of the cumulative hazard function (Figure \@ref(fig:separ)).

```{r separ, fig.cap = "Separation of exogenous and endogenous infant mortality (cumulative hazards, per cent).", fig.width = 5, fig.height = 4}
plot(fitall, fn = "cum", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
     lwd = 1.5, ylab = "Cumulative hazards")

axis(1, at = c(0, g(28), g(365)), labels = c(0, 28, 365), pos = 0)
axis(2, las = 1, pos = 0, at = c(0, 0.02, 0.04, 0.06), labels = c(0, 2, 4, 6))
#box()
abline(v = 0)
texhaz <- g(365) * fitex$hazards[1, 1]
lines(c(0, g(365)), c(0, texhaz), col = "red", lwd = 1.5)
atte <- c(0.0000, 0.0763, texhaz)
axis(4, at = atte, labels = 100 * round(atte, 4), las = 1, cex = 0.5,
     pos = g(365))
abline(v = g(365))
text(100, 0.03, "Exogenous", col = "red", srt = 22, cex = 0.8)
text(100, 0.053, "Endogenous plus exogenous", col = "blue", srt = 22, cex = 0.8)
axis(3, at =  c(0, g(28), g(365)), labels = c(0, 28, 365))
```

So we can see from Figure \@ref(fig:separ) that of the total mortality, 
`r round(5.27 / 7.63 * 100, 1)` per cent is exogenous and the rest is endogenous.
AS always, this requires that we trust the Bourgeois-Pichat model.

## Data

### Infant mortality, summary graphs

@@

```{r mortstat, fig.cap = "Crude infant mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
infdat <- readRDS("data/infdat3_year3.rds")
tid <- toTpch(Surv(enter, exit, event) ~ week, data = infdat, cuts = c(0, 1))
y <- tid$event / tid$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "IMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.03, 0.06, 0.09))
abline(h = c(0.05, 0.07, 0.09, 0.11), lty = 3)
abline(h = 0)
box()
```

@@

```{r neodata, fig.cap = "Crude neonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
neo <- age.window(infdat, c(0, 28 / 365))
tneo <- toTpch(Surv(enter, exit, event) ~ week, data = neo, cuts = c(0, 28 / 365))
y <- tneo$event / tneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "NMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, las = 1, at = c(0, 0.2, 0.4, 0.6))
abline(h = c(0.3, 0.4, 0.5, 0.6), lty = 3)
abline(h = 0)
box()
```

@@

```{r pneodata, fig.cap = "Crude postneonatal mortality by week of year, Umeå/Skellefteå 1895--1950.", fig.height = 4}
pneo <- age.window(infdat, c(28 / 365, 1))
tpneo <- toTpch(Surv(enter, exit, event) ~ week, data = pneo, cuts = c(28 / 365, 1))
y <- tpneo$event / tpneo$exposure
plot(1:52, y, type = "b", col = "blue", axes = FALSE, ylim = c(0, max(y)),
     xlab = "Week", ylab = "PNMR", cex = 0.7, lwd = 1.5)
axis(1, at = c(1, 10, 20, 30, 40, 52))
axis(2, at = c(0, 0.02, 0.04, 0.06), las = 1)
abline(h = c(0.03, 0.04, 0.05, 0.06, 0.07), lty = 3)
abline(h = 0)
box()
```

@@

```{r longtrend, fig.cap = "Crude IMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
trend <- toTpch(Surv(enter, exit, event) ~ year, data = infdat, cuts = c(0, 1))
y <- trend$event / trend$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "IMR", lwd = 1.5, cex = 0.7)
abline(h = c(0.05, 0.1, 0.15), lty = 3)
abline(h = 0)
```

```{r ntrend, fig.cap = "Crude NMR by year, Umeå-Skellefteå 1895--1950.", fig.height = 4}
ndat <- age.window(infdat, c(0, 28 / 365))
tnda <- toTpch(Surv(enter, exit, event) ~ year, data = ndat, cuts = c(0, 28 / 365))
y <- tnda$event / tnda$exposure
par(las = 1)
plot(1895:1951, y, type = "b", col = "blue", ylim = c(0, max(y)), xlab = "Year",
     ylab = "NMR", lwd = 1.5, cex = 0.7)
abline(h = 0)
abline(h = c(0.2, 0.4, 0.6), lty = 3)
```

### Temperature

@@

```{r monavgtemp, fig.cap="Weekly max and min temperature averages, 1895--1950.", fig.height = 4}
oldpar <- par(mfrow = c(1, 2))
## Umeå:
umaxavg <- umetemp$emaxtemp[1:52]
uminavg <- umetemp$emintemp[1:52]
plot(1:52, umaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Umeå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, uminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
## Skellefteå:
bmaxavg <- bjurtemp$emaxtemp[1:52]
bminavg <- bjurtemp$emintemp[1:52]
plot(1:52, bmaxavg, ylim = c(-20, 30), type = "b", col = "red", axes = FALSE,
     xlab = "Week", ylab = "Temperature (C)", cex = 0.6, lwd = 0.7)
text(17, 27, "Skellefteå")
axis(1, at = c(1, 10, 29, 43, 52))
axis(2, las = 1)
abline(h = 0)
abline(v = 29, lty = 3)
abline(h = c(-20, -10, 10, 20), lty = 3)
lines(1:52, bminavg, type = "b", col = "blue", cex = 0.6, lwd = 0.7)
text(8, 11, "max", col = "red")
text(20, -9, "min", col = "blue")
box()
##
par(oldpar)
```

Time trends of yearly average temperatures, see Figure

```{r longterm}
bmtemp <- rbind(umetemp, bjurtemp)
umt <- with(bmtemp, tapply(meantemp, year, mean))
plot(1894:1951, umt, type = "b", col = "blue", xlab = "Year",
     ylab = "Mean temperature", lwd = 1.5, cex = 0.7, ylim = c(0, 7))
abline(h = 0)
abline(h = 1:6, lty = 3)
```

## Statistical modelling
